---
title: "Specifying attribute hierarchies"
output: rmarkdown::html_vignette
bibliography: ../bib/references.bib
biblio-style: apa
csl: ../bib/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Specifying attribute hierarchies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| labels: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When specifying a diagnostic classification model (DCM), it is common to define relationships between the measured attributes.
For example, one attribute may be a precursor or successor to another attribute.
In other words, the attributes form a hierarchy.
In this article, we describe how to define different hierarchical relationships with dcmstan.

```{r}
#| label: load-dcmstan

library(dcmstan)
```

## Choosing a structural model

dcmstan supports two structural models that have the ability to specify an attribute hierarchy.
The hierarchical diagnostic classification model [HDCM\; @hdcm] can be defined with `hdcm()`.
The HDCM assumes a strict attribute hierarchy.
That is, possible profiles that conflict with the defined hierarchy are completely excluded from the model.
On the other hand, we can use a Bayesian Network [BayesNet\; @bayesnet] as the structural model using `bayesnet()`.
The BayesNet enforces a less restrictive hierarchy.
All possible profiles are still included in the model, but profiles that are inconsistent with the defined hierarchy are less likely.

Both `hdcm()` and `bayesnet()` include a `hierarchy` argument that can be used to define specific relationships between attributes.
In the following sections, we describe different attribute structures and show how to define these structures with dcmstan.

## Attribute structures

In general, hierarchies can be classified into two broad categories: simple and complex structures.


### Simple structures

For simple structures, attributes can exhibit a linear, converging, or diverging relationship.

```{r}
#| label: simple-structure
#| echo: false
#| message: false
#| fig-width: 8
#| fig-height: 3
#| fig-cap: Simple attribute structures

library(dagitty)
library(ggdag)
library(ggplot2)
library(dplyr)

dag_linear <- dagitty("dag { A1 -> A2 -> A3 }")
coordinates(dag_linear) <- list(
  x = c(A1 = 2, A2 = 2, A3 = 2),
  y = c(A1 = 2.6, A2 = 2, A3 = 1.4)
)

dag_converge <- dagitty("dag { A1 -> A3 <- A2 }")
coordinates(dag_converge) <- list(
  x = c(A1 = 1, A2 = 3, A3 = 2),
  y = c(A1 = 2.5, A2 = 2.5, A3 = 1.5)
)

dag_diverge <- dagitty("dag { A2 <- A1 -> A3 }")
coordinates(dag_diverge) <- list(
  x = c(A1 = 2, A2 = 1, A3 = 3),
  y = c(A1 = 2.5, A2 = 1.5, A3 = 1.5)
)

bind_rows(
  tidy_dagitty(dag_linear) |>
    mutate(type = "Linear") |>
    as_tibble(),
  tidy_dagitty(dag_converge) |>
    mutate(type = "Converging") |>
    as_tibble(),
  tidy_dagitty(dag_diverge) |>
    mutate(type = "Diverging") |>
    as_tibble()
) |>
  mutate(
    type = factor(type, levels = c("Linear", "Converging", "Diverging"))
  ) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  facet_wrap(~type, nrow = 1) +
  geom_dag_node() +
  geom_dag_text(color = "white") +
  geom_dag_edges(
    edge_color = "black",
    edge_width = 2,
    arrow_directed = grid::arrow(length = grid::unit(5, "pt"), type = "closed")
  ) +
  expand_limits(x = c(0, 4), y = c(1.3, 2.7)) +
  theme_void() +
  theme(
    strip.text = element_text(
      size = 12,
      margin = margin(0, 0, 4, 0),
      face = "bold"
    ),
    plot.margin = margin(5, 0, 0, 0)
  )
```

In a linear structure, the attributes must be acquired in a specific order, one after the other.
In the figure, a respondent must be proficient on A1 before they can acquire the skills in A2, followed by A3.
We can specify this relationship in dcmstan as:

```{r}
#| label: linear
#| results: hide

hdcm(hierarchy = "A1 -> A2 -> A3")
```

We use arrows (`->`) to denote dependencies between the attributes.
In the linear example, we defined the entire hierarchy with one continuous string of attributes (i.e., `"A1 -> A2 -> A3"`).
However, we could also define each pairwise relationship separately.
For example, `"A1 -> A2 A2 -> A3"` would be an equivalent specification.
Alternatively, we can also define the hierarchy in a separate variable that is passed to the structural model function.
Consider a converging structure, where proficiency of two or more attributes is required before a culminating attribute.
All of the following specifications are equivalent.

```{r}
#| label: converge
#| results: "hide"

hdcm(hierarchy = "A1 -> A3 <- A2")

hdcm(hierarchy = "A1 -> A3 A2 -> A3")

converge <- "
  A1 -> A3
  A2 -> A3
"
hdcm(hierarchy = converge)
```

Finally, we can similarly define a diverging structure, where one attribute serves as a precursor to two or more attributes.
Note that although we have used `hdcm()` in these examples, the same hierarchy specifications can be provided to `bayesnet()` as well.

```{r}
#| label: diverge
#| results: "hide"

hdcm(hierarchy = "A2 <- A1 -> A3")

hdcm(hierarchy = "A1 -> A2 A1 -> A3")

diverge <- "
  A1 -> A2
  A1 -> A3
"
hdcm(hierarchy = diverge)
```

### Complex structures

For complex structures, attributes can exhibit linear, converging, and diverging relationships within a single hierarchy. 
These configurations are defined by the types of simple structures present and how the structures are arranged in the hierarchy.

```{r}
#| label: complex-structure
#| echo: false
#| message: false
#| fig-width: 8
#| fig-height: 3.5
#| fig-cap: Complex attribute structures

dag_complex_2_level <- dagitty("dag { A3 <- A1 -> A4 <- A2 -> A5 }")
coordinates(dag_complex_2_level) <- list(
  x = c(A1 = 1.5, A2 = 2.5, A3 = 1, A4 = 2, A5 = 3),
  y = c(A1 = 2.4, A2 = 2.4, A3 = 1.6, A4 = 1.6, A5 = 1.6)
)

dag_complex_3_level <- dagitty("dag { A1 -> A2 -> A5 A1 -> A3 -> A5 
                                      A1 -> A4 -> A5 }")
coordinates(dag_complex_3_level) <- list(
  x = c(A1 = 2, A2 = 1, A3 = 2, A4 = 3, A5 = 2),
  y = c(A1 = 2.6, A2 = 2, A3 = 2, A4 = 2, A5 = 1.4)
)

dag_complex_4_level <- dagitty("dag { A1 -> A2 -> A3 -> A5 
                                      A1-> A2 -> A4 -> A5 }")
coordinates(dag_complex_4_level) <- list(
  x = c(A1 = 2, A2 = 2, A3 = 1, A4 = 3, A5 = 2),
  y = c(A1 = 2.6, A2 = 2.2, A3 = 1.8, A4 = 1.8, A5 = 1.4)
)

bind_rows(
  tidy_dagitty(dag_complex_2_level) |>
    mutate(type = "Two-Level Complex") |>
    as_tibble(),
  tidy_dagitty(dag_complex_3_level) |>
    mutate(type = "Three-Level Complex") |>
    as_tibble(),
  tidy_dagitty(dag_complex_4_level) |>
    mutate(type = "Four-Level Complex") |>
    as_tibble()
) |>
  mutate(
    type = factor(type, levels = c("Two-Level Complex", "Three-Level Complex",
                                   "Four-Level Complex"))
  ) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  facet_wrap(~type, nrow = 1) +
  geom_dag_node() +
  geom_dag_text(color = "white") +
  geom_dag_edges(
    edge_color = "black",
    edge_width = 2,
    arrow_directed = grid::arrow(length = grid::unit(5, "pt"), type = "closed")
  ) +
  expand_limits(x = c(0, 4), y = c(1.3, 2.7)) +
  theme_void() +
  theme(
    strip.text = element_text(
      size = 12,
      margin = margin(0, 0, 4, 0),
      face = "bold"
    ),
    plot.margin = margin(5, 0, 0, 0)
  )
```

Consider a two-level complex structure that includes two diverging structures and one converging structure. 
In this configuration, A1 and A2 serve as precursors to two attributes each.
A3 and A4 are the culminating attribute of A1, while A4 and A5 are the culminating attribute of A2.
A4 is the culminating attribute of both A1 and A2.

```{r}
#| label: "two-level-complex"
#| results: "hide"

hdcm(hierarchy = "A3 <- A1 -> A4 <- A2 -> A5")

hdcm(hierarchy = "A1 -> A3 A2 -> A5 A1 -> A4 <- A2")

complex2 <- "
  A1 -> A3
  A1 -> A4
  A2 -> A4
  A2 -> A5
"
hdcm(hierarchy = complex2)
```

Similarly, we can define a three-level complex structure that mixes linear, converging, and diverging structures.
In this configuration, A1 is a precursor to A2, A3, and A4. 
A5 is the culminating attribute of A2, A3, and A4.
This configuration forms three linear pathways from A1 to A5.


```{r}
#| label: "three-level-complex"
#| results: "hide"

hdcm(hierarchy = "A1 -> A2 -> A5 A1-> A3 -> A5 A1 -> A4 -> A5")

hdcm(hierarchy = "A1 -> A2 A2 -> A5 A1 -> A3 A3 -> A5 A1 -> A4 A4 -> A5")

complex3 <- "
  A1 -> A2
  A1 -> A3
  A1 -> A4
  A2 -> A5
  A3 -> A5
  A4 -> A5
"
hdcm(hierarchy = complex3)
```

Lastly, we can define a four-level complex structure that connects linear, diverging, and converging structures.
In this configuration, A1 is a precursor to A2.
A2 is a precursor to A3 and A4, and A5 is the culminating attribute of A3 and A4.

```{r}
#| label: "four-level-complex"
#| results: "hide"

hdcm(hierarchy = "A1 -> A2 -> A3 -> A5 A1 -> A2 -> A4 -> A5")

hdcm(hierarchy = "A1 -> A2 A2 -> A3 A3 -> A5 A2 -> A4 A4 -> A5")

complex4 <- "
  A1 -> A2
  A2 -> A3
  A2 -> A4
  A3 -> A5
  A4 -> A5
"
hdcm(hierarchy = complex4)
```

## References
